### 2014-04-02 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
figure script added

```{r, results='hide',  message=FALSE, error=FALSE}
includelibs = c("pheatmap", "sva", "qvalue", "limma")
tmp = lapply(includelibs, require, character.only=T)
print(tmp)
if(any(!unlist(tmp)))
{
  stop( paste("Not able to find all packages. Please install ",
              paste(includelibs[!unlist(tmp)], collapse=", ") )
              )
}
rm(tmp)
```
```{r, cache=TRUE, tidy=FALSE}
datamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))#[1:3000,]#dev.
sampleannotation = read.table("data/sampleInfoExample2.txt", 
                              sep="\t", header=TRUE, stringsAsFactors=FALSE)
rownames(sampleannotation)=sampleannotation$ArrayName
sampleannotation$Batch=factor(as.character(sampleannotation$Batch)) 
sampleannotation$Cell=factor(as.character(sampleannotation$Cell)) 
# must be discrete for the pheatmap
#table(sampleannotation$ArrayName==dimnames(datamatrix)[[2]])#ordercheck
datamatrix=datamatrix[,sampleannotation$Type!="WT"]
sampleannotation=sampleannotation[sampleannotation$Type!="WT",]
sampleannotation$Type=factor(sampleannotation$Type)
#dev/debug
useparprior=TRUE
print(dim(datamatrix))

```
Inspection of the covariate/batch balance;
```{r}
print(table(sampleannotation[ , c("Batch", "Type")])  )
```
Three batches and two covariates. There is a "Cell" column in the sample annotation, but it does not seem to have been used.

### Reproduce the original results
Following the description in section A.1 and A.2 and figure texts in the supplementary materials for Johnson et al., we try to reproduce some of their results. First, the heatmap in Figure A.1.
> A heatmap clustering of data set 2. 698 genes with large variation across all the samples are clustered.<cite> Johnson et al.

```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15, fig.cap="fig. 1", tidy=FALSE, results='hold', fig.show='hold'}
variationmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
clustermatrix = datamatrix[order(variationmeasure, decreasing=TRUE),][1:2698,]
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type")], 
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A1.",sep=""),border_color=NA,treeheight_row=0)
rm(variationmeasure, clustermatrix)
```
The heatmap does not look exactly as Fig. A1 in the paper. This could be due to undocumented transformations preformed in Johnson et al., for example log-transformation or standardization. And there are different ways of calculating variation for a probe. What parameters that were used for the clustering could also be different. But the purpose of the figure was to show that a batch effect is present in the data and especially for batch 3. This is also achieved in the reproduced figure.

Next 3 data sets were made,
- **EB2**: Batches 1 and 2 adjusted with non-parametric use of ComBat
- **EB3**: All thee batches adjusted with non-parametric use of ComBat
- **Batch3**: Only batch 3 (no adjustments)
```{r,results='hide', tidy=FALSE}
mat = datamatrix[,sampleannotation$Batch %in% c("1","2")]
EB2 = as.matrix( sva::ComBat(
            dat=mat, 
            batch=sampleannotation[colnames(mat),"Batch"], 
            mod=model.matrix( ~as.factor(sampleannotation[colnames(mat),"Type"])  ), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
rm(mat)
EB3 = as.matrix( sva::ComBat(
            dat=datamatrix, 
            batch=sampleannotation[colnames(datamatrix),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(datamatrix),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
Batch3 = datamatrix[,sampleannotation$Batch=="3"]

@@

The reproduced numbers are not the same as the reported ones, but they are not completely off. Again this is likely to be due to undocumented steps performed by Johnson et al.

@@

We start with the data before ComBat adjustment. Limma works best with log transformed and between array normalized values, so first we set negative values to 1 and filter out probes that has this in more than half the samples. To ease later comparisons the same filter will also be applied to the EB3 data set. Then the test is run with the batch included as a blocking factor.
```{r}
negativeprobesfilter =( rowSums(datamatrix>=1) >= (0.5*ncol(datamatrix)) ) & 
                      ( rowSums(EB3>=1) >= (0.5*ncol(EB3)) )
thismat = datamatrix[negativeprobesfilter, ]
thismat[thismat<1]=1
thismat=normalizeBetweenArrays(log2(thismat))
Type = as.factor(sampleannotation$Type)
Block = as.factor(sampleannotation$Batch)
design = model.matrix(~0+Type+Block)
fit = lmFit(thismat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
fit2 = contrasts.fit(fit, cont.matrix)
datamatrix_limma_res = eBayes(fit2)
rm(thismat, Type, Block, design, fit, cont.matrix, fit2)
print(table(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05))
```
Number of differentially expressed probes found with the alternative analysis is **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`**.

@@

Since limma was not the tool used in Johnson et al.'s workflow, we will first need to create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
```{r}
thismat = EB3[negativeprobesfilter, ]
thismat[thismat<1]=1
thismat=normalizeBetweenArrays(log2(thismat))
Type = as.factor(sampleannotation$Type)
design = model.matrix(~0 + Type)
fit = lmFit(thismat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
EB3_limma_res = eBayes(fit2)
rm(thismat, Type, design, fit, cont.matrix, fit2)
print(table(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05))

```
The number of genes below the significance threshold is fairly similar to the t-test.

Now we can compare the P-values for the two alternative procedures.
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EB3_limma_res$p.value[,1], border="blue", main="P-values, ComBat vs Limma", breaks=100, xlab="p-value")
hist(datamatrix_limma_res$p.value[,1], border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted"), text.col=c("blue", "red"))
```
This plot shows that the p-values calculated from the ComBat adjusted data is skewed towards the low end compared to when batch is considered inside the statistical test in limma. However, the difference is not large and from this plot alone it could be argued that ComBat is just better.

Lets assume one aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05. Then handling the batch effect with ComBat adjustment will yield a list of **`r sum(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05)`** whereas a list of **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`** probes would have been closer to the truth. Also it is of interest to note that it is mostly the same probes that are found.
```{r, tidy=FALSE}
table( qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05 ,
         qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05,
       dnn=c("limma", "ComBat"))
```
And if they were to use the top 1000 probes in a gene set test, they would have found many of the same genes regardless of handling batch effects by ComBat or limma.
```{r, tidy=FALSE}
table( rank(datamatrix_limma_res$p.value[,1])  <=1000 & 
         rank(EB3_limma_res$p.value[,1])<=1000   )
```
Our conclusion is that for this study the error introduced by the use of ComBat would probably have a modest effect on the final result.

@@

First we use random numbers drawn from the same distribution regardless of batch or covariate but retaining the batch/covariate design.
```{r, results='hide', tidy=FALSE}
set.seed(100)
randommatrix = datamatrix
randommatrix[,] =rnorm(length(datamatrix), mean=0, sd=1)
EBrand = as.matrix( sva::ComBat(
            dat=randommatrix, 
            batch=sampleannotation[colnames(randommatrix),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(randommatrix),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
```
Limma is then used in 3 ways
- On the ComBat adjusted random numbers
- On the random numbers including batch as a blocking factor
- On the random numbers ignoring batch information
```{r, tidy=FALSE}

@@

fit = lmFit(EBrand, design)
fit2 = contrasts.fit(fit, cont.matrix)
EBrand_limma_pvalues = eBayes(fit2)$p.value[,1]

fit = lmFit(randommatrix, design)
fit2 = contrasts.fit(fit, cont.matrix)
randommatrix_limma_pvalues = eBayes(fit2)$p.value[,1]

Block = as.factor(sampleannotation$Batch)
design = model.matrix(~0+Type+Block)
fit = lmFit(randommatrix, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design) 
fit2 = contrasts.fit(fit, cont.matrix)
fit2 = eBayes(fit2)
randommatrix_limma_batch_pvalues = fit2$p.value[,1]

```
Plotting the p-values
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EBrand_limma_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
hist(randommatrix_limma_pvalues, border="black", add=T, breaks=100)
hist(randommatrix_limma_batch_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=c("blue", "black", "red"))
```
The p-values for ComBat adjusted random numbers are slightly enriched for low p-values. Indicating that some of the enrichment seen in the real data is also false.

Another sanity check is to subset the real data into fictive batches. The idea behind this check stems from the observation that the EB2 data set contains the same number of samples with the same batch/covariate balance as the third batch alone. 
> The third batch was used for comparison against the EB2 analysis results because it was an identical experiment to EB2 other than the fact that it was conducted in a single batch. <cite> Johnson et al.

Thus it is interesting to inspect the p-value from these two identical experiments, EB2 (batch 1 and 2 adjusted with ComBat) and batch 3 (no ComBat adjustment). These p-values are already computed above.
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EB2_pvals, border="blue",  main="P-values, real data", breaks=100, xlab="p-value")
hist(Batch3_pvals, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted batch 1 and 2", "Batch 3"), text.col=c("blue", "red"))
```
The 2-batch experiment is able to retrieve more genes as significant than running all in one batch for the same sample size. This is also observed by Johnson et al., but not commented. We claim that the apparent better result for the two-batch experiment is likely a consequence of the use of ComBat. A test of this is to split batch 3 into two imaginary batches with the same design as for batch 1 and 2, and then look at the p-value distribution compared with the real batch 3.

```{r, tidy=FALSE}
Batch45 = Batch3
Batch45_annot = sampleannotation[dimnames(Batch45)[[2]],]
Batch45_annot$Batch="4"
Batch45_annot$Batch[Batch45_annot$Type=="C"][1:4] = "5"
Batch45_annot$Batch[Batch45_annot$Type=="R"][1:3] = "5"
print(table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")]))
print(table(Batch45_annot[, c("Batch", "Type")]))
```
Then adjust the data with fictive batches and calculate p-values as done for the un-adjusted batch 3.
```{r, results='hide', tidy=FALSE}
EB45 = as.matrix( sva::ComBat(
            dat=Batch45, 
            batch=Batch45_annot[colnames(Batch45),"Batch"], 
            mod=model.matrix( ~as.factor(Batch45_annot[colnames(Batch45),"Type"])  ), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))

EB45_pvals = apply(EB45, 1 , 
                  FUN=function(x){t.test(
                                    x[Batch45_annot[colnames(EB45), "Type"]=="C"],
                                    x[Batch45_annot[colnames(EB45), "Type"]=="R"]
                                    )$p.value})
```
```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
hist(EB45_pvals, border="blue",  main="P-values, real data", breaks=100, xlab="p-value")
hist(Batch3_pvals, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted fictive-batches of batch 3", "Batch 3 not adjusted"), text.col=c("blue", "red"))
```
The fictive-batches version with ComBat seemingly outperforms the original. This plot looks very similar to the EB2 vs. Batch3 p-values, and for both, the p-values for the ComBat adjusted data are likely deflated due to ComBat use.

The fictive batch asssignment above was just one of several possible. Repeating the selection 10 times shows that the lower p-values was not a coincidence.

```{r, tidy=FALSE, results='hide'}
fictivebatches_pvalcounts=list()
runs = 10
for(i in 1:runs)
{
Batch45 = Batch3
Batch45_annot = sampleannotation[dimnames(Batch45)[[2]],]
Batch45_annot$Batch="4"
Batch45_annot$Batch[Batch45_annot$Type=="C"][sample(1:sum(Batch45_annot$Type=="C"), 4)] = "5"
Batch45_annot$Batch[Batch45_annot$Type=="R"][sample(1:sum(Batch45_annot$Type=="R"), 3)] = "5"
print(table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")]))
print(table(Batch45_annot[, c("Batch", "Type")]))

EB45 = as.matrix( sva::ComBat(
            dat=Batch45, 
            batch=Batch45_annot[colnames(Batch45),"Batch"], 
            mod=model.matrix( ~as.factor(Batch45_annot[colnames(Batch45),"Type"])  ), 
            numCovs=NULL, 
            par.prior=TRUE, 
            prior.plots=FALSE))

pvals = apply(EB45, 1 , 
                  FUN=function(x){t.test(
                                    x[Batch45_annot[colnames(EB45), "Type"]=="C"],
                                    x[Batch45_annot[colnames(EB45), "Type"]=="R"]
                                    )$p.value})

fictivebatches_pvalcounts[[i]] = hist(pvals, plot=FALSE, breaks=100)$counts
}
```
In order to save some computational time the parameteric version of ComBat was used for this.
```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
plot((1:20)/100, hist(Batch3_pvals, plot=FALSE, breaks=100)$counts[1:20], col="red",  main="P-values, fictive batches", 
      xlab="p-value", ylab="Frequency", type="l", lwd=2, 
      ylim=c(0, max(unlist(fictivebatches_pvalcounts))))
for(i in 1:length(fictivebatches_pvalcounts))
{
  lines((1:20)/100,fictivebatches_pvalcounts[[i]][1:20], col="blue")
}
legend("topright", legend=c("ComBat adjusted fictive-batches of batch 3", "Batch 3 not adjusted"), text.col=c("blue", "red"))
```

```{r}
source("../../commonscripts/helper_functions.r")
nshuffleddatasets = 10
datamatrices_permuted = list()
for(i in 1:nshuffleddatasets)
{
  x = shufflesamplesinbatch( rownames(sampleannotation), sampleannotation$Batch, sampleannotation$Type, c("C", "R") )
  print(  sum(sampleannotation[,"Type"] != sampleannotation[x,"Type"]  ))  # see how many changed
  datamatrices_permuted[[i]] = as.matrix(ComBat(dat=datamatrix[, x], 
                                                batch=sampleannotation$Batch, 
                                                mod=model.matrix(~as.factor(sampleannotation[colnames(datamatrix),"Type"])),  
                                                numCovs=NULL, 
                                                par.prior=TRUE, 
                                                prior.plots=FALSE))
}

permuted_pvalcounts=list()
for(i in 1:nshuffleddatasets)
{ 
  this_p = apply(datamatrices_permuted[[i]], 1 , 
                  FUN=function(x){t.test(
                                    x[sampleannotation[, "Type"]=="C"],
                                    x[sampleannotation[, "Type"]=="R"]
                                    )$p.value})

  permuted_pvalcounts[[i]] = hist(this_p, plot=FALSE, breaks=100)$counts
  print(table(p.adjust(this_p, method="fdr")<0.05))
}
```

@@

# reproduced ComBat + limma
hist(EB3_limma_res$p.value[,1], main="P-values, real data, ComBat followed by Limma", breaks=100, xlab="p-value", ylim=c(0,3000))

# batch handled in limma
hist(datamatrix_limma_res$p.value[,1], breaks=100, main="P-values, real data, batch handled by Limma", xlab="p-value" , ylim=c(0,3000))

>>>>

```{r, results='hide',  message=FALSE, error=FALSE}
includelibs = c("pheatmap", "sva", "qvalue", "limma")
tmp = lapply(includelibs, require, character.only=T)
print(tmp)
if(any(!unlist(tmp)))
{
  stop( paste("Not able to find all packages. Please install ",
              paste(includelibs[!unlist(tmp)], collapse=", ") )
              )
}
rm(tmp)
debug=FALSE
```
```{r loaddata,  tidy=FALSE}
datamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))
if(debug)datamatrix = datamatrix[1:3000,]
sampleannotation = read.table("data/sampleInfoExample2.txt", 
                              sep="\t", header=TRUE, stringsAsFactors=FALSE)
rownames(sampleannotation)=sampleannotation$ArrayName
sampleannotation$Batch=factor(as.character(sampleannotation$Batch)) 
sampleannotation$Cell=factor(as.character(sampleannotation$Cell)) 
# must be discrete for the pheatmap
#table(sampleannotation$ArrayName==dimnames(datamatrix)[[2]])#ordercheck
datamatrix=datamatrix[,sampleannotation$Type!="WT"]
sampleannotation=sampleannotation[sampleannotation$Type!="WT",]
sampleannotation$Type=factor(sampleannotation$Type)
#dev/debug
useparprior=TRUE
print(dim(datamatrix))

```
Inspection of the covariate/batch balance;
```{r}
print(table(sampleannotation[ , c("Batch", "Type")])  )
```
Three batches and two covariates. There is a "Cell" column in the sample annotation, but it does not seem to have been used.
<br/>  
<br/>  
### Reproduce the original results
Following the description in section A.1 and A.2 and figure texts in the supplementary materials for Johnson et al., we try to reproduce some of their results. First, the heatmap in Figure A.1.
> A heatmap clustering of data set 2. 698 genes with large variation across all the samples are clustered.<cite> Johnson et al.

```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15, fig.cap="fig. 1", tidy=FALSE, results='hold', fig.show='hold'}
variationmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
clustermatrix = datamatrix[order(variationmeasure, decreasing=TRUE),][1:2698,]
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type")], 
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A1.",sep=""),border_color=NA,treeheight_row=0)
rm(variationmeasure, clustermatrix)
```
The heatmap does not look exactly as [Fig. A.1](http://biostatistics.oxfordjournals.org/content/suppl/2006/04/21/kxj037.DC1/kxj037supp.pdf) in Johnson et al. This could be due to undocumented transformations preformed in Johnson et al., for example log-transformation or standardization. And there are different ways of calculating variation for a probe. The parameters used in the clustering could also be different. But the purpose of the figure was to show that a batch effect is present in the data and especially for batch 3. This is also achieved in the reproduced figure.
<BR/>  
<BR/>  
Next 3 data sets were made,
- **EB2**: Batches 1 and 2 adjusted with ComBat
- **EB3**: All the batches adjusted with ComBat
- **Batch3**: Only batch 3 (no adjustments)
```{r,results='hide', tidy=FALSE}
mat = datamatrix[,sampleannotation$Batch %in% c("1","2")]
EB2 = as.matrix( sva::ComBat(
            dat=mat, 
            batch=sampleannotation[colnames(mat),"Batch"], 
            mod=model.matrix( ~as.factor(sampleannotation[colnames(mat),"Type"])  ), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
rm(mat)
EB3 = as.matrix( sva::ComBat(
            dat=datamatrix, 
            batch=sampleannotation[colnames(datamatrix),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(datamatrix),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
Batch3 = datamatrix[,sampleannotation$Batch=="3"]

@@

The reproduced numbers are not the same as the reported ones, but they are not completely off. This is likely due to undocumented steps performed by Johnson et al.

The above effort done in reproducing some of the results for data set 2 is done using non-log transformed data that also includes some negative values. In the next part of this report we will do an alternative analysis were batch is included in the analysis using limma instead of using ComBat. Since Limma needs log2 transformed data, the results will not be directly comparable to a t-test result using non-log values. Later we will also perform a few test using permutation (i.e many runs). For these reasons the rest of the analysis, including the significance test on ComBat adjusted data, will be performed on log2 transformed data with negative values set to 1 or removed (if in many samples for that probe). This is a fairly standard workflow and more practical for doing comparisons. In addition the parametric version of ComBat will be used.

First we make a data matrix with log2 values. And doing so the negative values will be floored and removed if in 
```{r}
log2data = datamatrix

# flooring to 1
log2data[log2data<1]=1

# take out data with to much low/missing values.
negativeprobesfilter =( rowSums(log2data>1) >= (0.9*ncol(log2data)) )
log2data = log2data[negativeprobesfilter,]

# quantilenormalize
log2data=normalizeBetweenArrays(log2(log2data))
```

```{r}

log2dataComBat = as.matrix( sva::ComBat(
            dat=log2data, 
            batch=sampleannotation[colnames(log2data),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(log2data),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))

Type = as.factor(sampleannotation$Type)
design = model.matrix(~0 + Type)
fit = lmFit(log2dataComBat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
log2ComBat_res = eBayes(fit2)
rm(Type, design, fit, cont.matrix, fit2)
print(table(qvalue(log2ComBat_res$p.value[,1])$qvalue<0.05))

```
The number of genes below the significance threshold is somewhat lower than for the original the t-test, but that comparison is not relevant since the data is processed in different ways before the test. But for the explained practical purposes this is the result coming from the ComBat adjusted data.

@@

We start with the data before ComBat adjustment. Limma works best with log transformed and between array normalized values, so first we set negative values to 1 and filter out probes that has this in more than half the samples. To ease later comparisons the same filter will also be applied to the EB3 data set. Then the test is run with the batch included as a blocking factor.
```{r}
Type = as.factor(sampleannotation$Type)
Block = as.factor(sampleannotation$Batch)
design = model.matrix(~0+Type+Block)
fit = lmFit(log2data, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
fit2 = contrasts.fit(fit, cont.matrix)
log2limmablock_res = eBayes(fit2)
rm(Type, Block, design, fit, cont.matrix, fit2)
print(table(qvalue(log2limmablock_res$p.value[,1])$qvalue<0.05))
```
Number of differentially expressed probes found with the alternative analysis is **`r sum(qvalue(log2limmablock_res$p.value[,1])$qvalue<0.05)`**.

@@

Now we can compare the P-values for the two alternative procedures.
```{r, dev='svg', fig.width=8, fig.height=8}
hist(log2ComBat_res$p.value[,1], border="blue", main="P-values, ComBat vs Limma", breaks=100, xlab="p-value")
hist(log2limmablock_res$p.value[,1], border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted"), text.col=c("blue", "red"))
```
This plot shows that the p-values calculated from the ComBat adjusted data is skewed towards the low end compared to when batch is considered inside the statistical test in limma. However, the difference is not large and from this plot alone it could be argued that ComBat is just better.

Lets assume one aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05. Then handling the batch effect with ComBat adjustment will yield a list of **`r sum(qvalue(log2ComBat_res$p.value[,1])$qvalue<0.05)`** whereas a list of **`r sum(qvalue(log2limmablock_res$p.value[,1])$qvalue<0.05)`** probes would have been closer to the truth. Also it is of interest to note that it is mostly the same probes that are found.
```{r, tidy=FALSE}
table( qvalue(log2limmablock_res$p.value[,1])$qvalue<0.05 ,
         qvalue(log2ComBat_res$p.value[,1])$qvalue<0.05,
       dnn=c("limma", "ComBat"))
```
And if they were to use the top 1000 probes in a gene set test, they would have found many of the same genes regardless of handling batch effects by ComBat or limma.
```{r, tidy=FALSE}
table( rank(log2limmablock_res$p.value[,1])  <=1000 & 
         rank(log2ComBat_res$p.value[,1])<=1000   )
```
Our conclusion is that for this study the error introduced by the use of ComBat would probably have a modest effect on the final result.

@@

First we use random numbers drawn from the same distribution regardless of batch or covariate but retaining the batch/covariate design.
```{r, results='hide', tidy=FALSE}
set.seed(100)
randomdata = log2data
randomdata[,] =rnorm(length(randomdata), mean=0, sd=1)
randomdataComBat = as.matrix( sva::ComBat(
            dat=randomdata, 
            batch=sampleannotation[colnames(randomdata),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(randomdata),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
```
Limma is then used in 3 ways
- On the ComBat adjusted random numbers
- On the random numbers including batch as a blocking factor
- On the random numbers ignoring batch information
```{r, tidy=FALSE}

@@

fit = lmFit(randomdataComBat, design)
fit2 = contrasts.fit(fit, cont.matrix)
randomdataComBat_pvalues = eBayes(fit2)$p.value[,1]

fit = lmFit(randomdata, design)
fit2 = contrasts.fit(fit, cont.matrix)
randomdata_pvalues = eBayes(fit2)$p.value[,1]

Block = as.factor(sampleannotation$Batch)
design = model.matrix(~0+Type+Block)
fit = lmFit(randomdata, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design) 
fit2 = contrasts.fit(fit, cont.matrix)
fit2 = eBayes(fit2)
randomdatalimmablock_pvalues = fit2$p.value[,1]

```
Plotting the p-values
```{r, dev='svg', fig.width=8, fig.height=8}
hist(randomdataComBat_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
hist(randomdata_pvalues, border="black", add=T, breaks=100)
hist(randomdatalimmablock_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=c("blue", "black", "red"))
```
The p-values for ComBat adjusted random numbers are slightly enriched for low p-values. Indicating that some of the enrichment seen in the real data is also false.
<br/>  
<br/>  
Another sanity check is to subset the real data into fictive batches. The idea behind this check stems from the observation that the EB2 data set contains the same number of samples with the same batch/covariate balance as the third batch alone.

> The third batch was used for comparison against the EB2 analysis results because it was an identical experiment to EB2 other than the fact that it was conducted in a single batch. <cite> Johnson et al.

Thus it is interesting to inspect the p-value from these two identical experiments, EB2 (batch 1 and 2 adjusted with ComBat) and batch 3 (no ComBat adjustment). These p-values are already computed above. 
```{r, dev='svg', fig.width=8, fig.height=8}
# need to subset the log2data and do the ComBat adjustment and do the limma tests

# subset log2data
mat = log2data[,sampleannotation$Batch %in% c("1","2")]
log2EB2 = as.matrix( sva::ComBat(
            dat=mat, 
            batch=sampleannotation[colnames(mat),"Batch"], 
            mod=model.matrix( ~as.factor(sampleannotation[colnames(mat),"Type"])  ), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
rm(mat)
log2Batch3 = log2data[,sampleannotation$Batch=="3"]

# run limma
Type = as.factor(sampleannotation[colnames(log2EB2), "Type"])
design = model.matrix(~0 + Type)
fit = lmFit(log2EB2, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
log2EB2ComBat_pvals = eBayes(fit2)$p.value
rm(Type, design, fit, cont.matrix, fit2)

Type = as.factor(sampleannotation[colnames(log2Batch3), "Type"])
design = model.matrix(~0 + Type)
fit = lmFit(log2Batch3, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
log2Batch3ComBat_pvals = eBayes(fit2)$p.value
rm(Type, design, fit, cont.matrix, fit2)

# plot the two p-value distibutions
hist(log2EB2ComBat_pvals, border="blue",  main="P-values, real data", breaks=100, xlab="p-value")
hist(log2Batch3ComBat_pvals, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted batch 1 and 2", "Batch 3"), text.col=c("blue", "red"))
```
The 2-batch experiment is able to retrieve more genes as significant(`r sum(qvalue(log2EB2ComBat_pvals)$qvalue<0.05)` for q<0.05) than running all in one batch(`r sum(qvalue(log2Batch3ComBat_pvals)$qvalue<0.05)` for q<0.05) for the same sample size. This is also observed by Johnson et al., but not commented. We claim that the apparent better result for the two-batch experiment is a consequence of the use of ComBat. A sanity check supporting this claim is to split batch 3 into two imaginary batches with the same design as for batch 1 and 2, and then look at the p-value distribution compared with the real batch 3. In order to obtain some robustness this test in performed for 10 permutations of different ways of sub-setting batch 3 into 2 fictive batches.

```{r, tidy=FALSE, results='hide'}

# print the batch/covariate balance for the real batch 1 and batch 2
print(table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")]))

fictivebatches_pvalcounts=list()
runs = 10
for(i in 1:runs)
{
permdata = log2Batch3
permdata_annot = sampleannotation[dimnames(permdata)[[2]],]
permdata_annot$Batch="4"
permdata_annot$Batch[permdata_annot$Type=="C"][sample(1:sum(permdata_annot$Type=="C"), 4)] = "5"
permdata_annot$Batch[permdata_annot$Type=="R"][sample(1:sum(permdata_annot$Type=="R"), 3)] = "5"

# This permutations balance
print(table(permdata_annot[, c("Batch", "Type")]))

permdataComBat = as.matrix( sva::ComBat(
            dat=permdata, 
            batch=permdata_annot[,"Batch"], 
            mod=model.matrix( ~as.factor(permdata_annot[,"Type"])  ), 
            numCovs=NULL, 
            par.prior=TRUE, 
            prior.plots=FALSE))

Type = as.factor(permdata_annot[, "Type"])
design = model.matrix(~0 + Type)
fit = lmFit(permdataComBat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
pvals = eBayes(fit2)$p.value

fictivebatches_pvalcounts[[i]] = hist(pvals, plot=FALSE, breaks=100)$counts
}
```

The p-values for the 10 splits of Batch 3 adjusted by ComBat is compared to the p-values for the not-split(and not ComBat adjusted) Batch 3.
```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
plot((1:20)/100, hist(log2Batch3ComBat_pvals, plot=FALSE, breaks=100)$counts[1:20],
     col="red",  main="P-values, fictive batches", 
      xlab="p-value", ylab="Frequency", type="l", lwd=2, 
      ylim=c(0, max(unlist(fictivebatches_pvalcounts))))
for(i in 1:length(fictivebatches_pvalcounts))
{
  lines((1:20)/100,fictivebatches_pvalcounts[[i]][1:20], col="blue")
}
legend("topright", legend=c("ComBat adjusted fictive-batches of batch 3", "Batch 3 not adjusted"), text.col=c("blue", "red"))
```
The two-batch approach seem to consistently outperform the all-in-one-batch version.

Alternativ sanitycheck som data set 2 ikke feiler paa, saa jeg synes ikke den skal med. Laget for aa vise i figur som eksempel
```{r}
source("../../commonscripts/helper_functions.r")
nshuffleddatasets = 10
datamatrices_permuted = list()
for(i in 1:nshuffleddatasets)
{
  permlabels = shufflesamplesinbatch( rownames(sampleannotation), sampleannotation$Batch, sampleannotation$Type, c("C", "R") )
  print(  sum(sampleannotation[,"Type"] != sampleannotation[permlabels,"Type"]  ))  # see how many changed
  datamatrices_permuted[[i]] = as.matrix(ComBat(dat=datamatrix[, permlabels], 
                                                batch=sampleannotation$Batch, 
                                                mod=model.matrix(~as.factor(sampleannotation[colnames(datamatrix),"Type"])),  
                                                numCovs=NULL, 
                                                par.prior=TRUE, 
                                                prior.plots=FALSE))
}

permuted_pvalcounts=list()
for(i in 1:nshuffleddatasets)
{

  Type = as.factor(sampleannotation$Type)
  design = model.matrix(~0 + Type)
  fit = lmFit(datamatrices_permuted[[i]], design)
  cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
  fit2 = contrasts.fit(fit, cont.matrix)
  this_p = eBayes(fit2)$p.value[,1]

  permuted_pvalcounts[[i]] = hist(this_p, plot=FALSE, breaks=100)$counts
  print(table(qvalue(this_p)$qvalue<0.05))
}
```

@@

# real data, reproduced ComBat + limma
hist(log2ComBat_res$p.value[,1], main="P-values, real data, ComBat followed by Limma", breaks=100, xlab="p-value", ylim=c(0,3000))

# real data, batch handled in limma
hist(log2limmablock_res$p.value[,1], breaks=100, main="P-values, real data, batch handled by Limma", xlab="p-value" , ylim=c(0,3000))

# random ComBat + limma
hist(randomdataComBat_pvalues, main="P-values, Random data, ComBat followed by Limma", breaks=100, xlab="p-value" , ylim=c(0,3000))
